Error Reporting Parsers: a Monad Transformer Approach

### 2013-08-07 article.tex
Final copy
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

The language definition of Woof in pseudo-BNF~\cite{bnf} is:

@@

\subsection{Type Class Preliminaries}
Unfortunately, the standard MonadError instance for the \verb+Either+ datatype does
not fit our needs -- it requires an \verb+Error+ constraint that we do not want.
We'll implement our own MonadError type class, based on the
standard one, and instances of it.  We'll need an instance for each member of
our stack -- Either, StateT, and MaybeT: 
\begin{verbatim}
class Monad m => MonadError e m | m -> e where
  throwError :: e -> m a
  catchError :: m a -> (e -> m a) -> m a

instance MonadError e (Either e) where
  throwError               =  Left
  catchError  (Right x) _  =  Right x
  catchError  (Left e)  f  =  f e

instance MonadError e m => MonadError e (StateT s m) where
  throwError      =  lift . throwError
  catchError m f  =  StateT g
    where
      g s = catchError (runStateT m s) 
                       (\e -> runStateT (f e) s)

instance MonadError e m => MonadError e (MaybeT m) where
  throwError      =  lift . throwError
  catchError m f  =  MaybeT $ catchError (runMaybeT m) 
                                         (runMaybeT . f)
\end{verbatim}
Note how the StateT and MaybeT instances don't really do anything with the
errors; they just pass the error on through to the next level.  That's why they
both require that the next lower level supports MonadError.  Also, this code 
requires some additional extensions:
\begin{itemize}
  \item FlexibleInstances
  \item FunctionalDependencies
  \item UndecidableInstances 
\end{itemize}

@@

The MonadError type class provides two useful combinators: \verb+throwError+ for 
generating errors and \verb+catchError+ for dealing with them.  We can ignore 
the second because we don't need to recover from errors, just report them, 
but when should an error be generated and what 
data should it include?  \verb+throwError+ is too basic for our needs, so we'll 
introduce the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError e m, Alternative m) => e -> m a -> m a
commit err p = p <|> throwError err
\end{verbatim}

@@

And so on.

@@

Now that we have two separate state layers, we need to be sure not to mix them 
up (fortunately, Haskell's type system will helpfully complain if I mess up);
the \verb+lift+ method of the \verb+MonadTrans+ type class allows us to access
the position state:
\begin{verbatim}
getState :: Parser Char Pos
getState = lift get

@@

\Red{It's worth noting that up until this point, 
we haven't talked about how transformers contribute to the design of the parser library.  
Maybe give the juicy bits earlier.}

@@

We benefitted from these advantages when many of the parsers didn't have to change 
at all from one version to the next (although some of their types did change).

@@

\section{Conclusion}
Monad transformers provide an elegant solution to error reporting in parser
combinators.  Better yet, they are extensible -- we could go back and add more
features to our parsers using additional transformers, perhaps extending our
parser to save partial results, so that if there's an error, it reports the
progress that it had made before the error.  We could do that using a Writer
transformer.  Or we check to make sure that all variables are in scope when
they're used -- using a Reader transformer.  We could even capture whitespace
and comment tokens, instead of just throwing them away, in case they contained
valuable information.  And the best part of it is, most of the parsers we already
wrote would continue to work fine without needing any modifications.  That
is the power of monad transformers!

>>>>

The language definition of Woof in pseudo-BNF is:

@@

\subsection{Type Class Preliminaries}
Unfortunately, the standard MonadError instance for the \verb+Either+ datatype does
not fit our needs -- it requires an \verb+Error+ constraint that we do not want.
We'll implement our own MonadError type class.  We'll need an instance for each member of
our stack -- Either, StateT, and MaybeT:
\begin{verbatim}
{-# LANGUAGE FlexibleContexts, MultiParamTypeClasses,
    FlexibleInstances, TypeFamilies #-}

class Monad m => MonadError m where
  type Error m :: *
  throwError :: Error m -> m a
  catchError :: m a -> (Error m -> m a) -> m a

instance MonadError (Either e) where
  type Error (Either e) = e
  throwError               =  Left
  catchError  (Right x) _  =  Right x
  catchError  (Left e)  f  =  f e

instance MonadError m => MonadError (StateT s m) where
  type Error (StateT s m) = Error m
  throwError      =  lift . throwError
  catchError m f  =  StateT g
    where
      g s = catchError (runStateT m s) 
                       (\e -> runStateT (f e) s)

instance MonadError m => MonadError (MaybeT m) where
  type Error (MaybeT m) = Error m
  throwError      =  lift . throwError
  catchError m f  =  MaybeT $ catchError (runMaybeT m)
                                         (runMaybeT . f)
\end{verbatim}
Note how the StateT and MaybeT instances don't really do anything with the
errors; they just pass the error on through to the next level.  That's why they
both require that the next lower level supports MonadError.  This code uses
type families to state what the error type is; explaining type families
in more detail is out of scope for this article, but roughly, given some MonadError \verb|m|,
you can write down the type of error that monad supports by saying \verb|Error m|.

@@

The MonadError type class provides two useful combinators: \verb+throwError+ for 
generating errors and \verb+catchError+ for dealing with them.  We can ignore 
the second because we don't need to recover from errors, just report them, 
but when should an error be generated and what 
data should it include?  \verb+throwError+ will unconditionally throw an
error, but often we will only want to conditionally throw an error. Thus, we
introduce the \verb+commit+ combinator:
\begin{verbatim}
commit :: (MonadError m, Alternative m) => Error m -> m a -> m a
commit err p = p <|> throwError err
\end{verbatim}

@@

@@

Now that we have two separate state layers, we need to be sure not to mix them 
up (fortunately, because the types of the states are different, Haskell's type system will helpfully complain if I mess up).
The \verb+lift+ method of the \verb+MonadTrans+ type class allows us to access
the position state:
\begin{verbatim}
getState :: Parser Char Pos
getState = lift get

@@

@@

We could see these advantages when many of the parsers didn't have to change 
at all from one version to the next (although some of their types did change).

@@

\section{Conclusion}
Monad transformers provide an elegant solution to error reporting in parser
combinators.  Better yet, they are extensible -- we could go back and add more
features to our parsers using additional transformers, perhaps extending our
parser to save partial results, so that if there's an error, it reports the
progress that it had made before the error.  We could do that using a Writer
transformer.  Or we check to make sure that all variables are in scope when
they're used using a Reader transformer.  We could even capture whitespace
and comment tokens, instead of just throwing them away, in case they contained
valuable information.  And the best part of it is, most of the parsers we already
wrote would continue to work fine without needing any modifications.  That
is the power of monad transformers!
